Dissertation Proposal

### 2023-11-10 proposal.tex
grammar check

% TODO
% - clean up the background sections

@@

%% abbrev. for referring to supplementary materials
\newcommand{\supplRef}[1]{\texttt{#1} in the supplementary materials}

@@

  I propose a plan for finding misconceptions possibly not anticipated
  by experts, and the design of an automated, self-guided tutoring
  system. The tutor builds on strategies in the cognitive and
  educational literature and is explicitly designed around identifying
  and correcting misconceptions. Preliminary results suggest (a) the
  misconceptions I found are widespread, and (b) the tutor appears to
  improve understanding.
\end{abstract}

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions (e.g.,
Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python
\citep{bernhardtWat2012, guhaEssenceJavascript2010,
politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still,
this semantic core bridges many syntaxes, and understanding it helps
when transferring knowledge from old languages to new ones. In
recognition of this deep commonality, in this paper I choose to call
this the \emph{Standard Model of Languages} (SMoL).

@@

This lead to my \textbf{research questions}:

@@

% To answer my research questions, I have (a) run a multi-year,
% multi-phase study to identify these misunderstandings while trying to
% work around the blind spots of experts (and also drawing on prior
% work), and (b) constructed the SMoL Tutor, an interactive tutor to
% address these misunderstandings. The SMoL Tutor is not a
% \emph{programming} tutor; rather, it assumes the user has some basic
% facility with programming (i.e., that they are already familiar with
% concepts like vectors, mutation, and higher-order functions). It is
% built around SMoL's features and their interactions, and draws on
% cognitive and educational psychology concepts to expressly identify
% and correct misconceptions.

% Concretely, I have made the following contributions:
% \begin{enumerate}

%   \item I provide a collection of brief programs for detecting
%         misconceptions about the features of SMoL and their
%         interaction. These programs are short and most are readily
%         translatable to a wide variety of programming languages that
%         share the SMoL characteristics.

%   \item I show that multiple populations have problems with these
%         programs.

%   \item I present a list of curated misconceptions generated
%         after multiple rounds of analysis.

%   \item I implement these misconceptions as interpreters, which
%         find weaknesses in my manual analysis and hold potential for
%         generative use in the future.

%   \item I present a tutoring system, the SMoL Tutor, that measurably
%         corrects these misconceptions.

%   \item I present a notional machine, the Stacker, that is used by the
%         SMoL Tutor to explain programs and might be useful in itsown.

% \end{enumerate}

% This document first presents the current results and ends with future
% work \& timeline (\cref{s:future-work}).

% \paragraph{A Note on Terminology} I use the term
% ``behavior'' to refer to the meaning of programs in terms of the
% answers they produce. A more standard term for this would, of course, be
% ``semantics''. However, the term ``semantics tutor'' might
% mislead some readers into thinking it
% teaches people to read or write a formal semantics, e.g., an
% introduction to ``Greek'' notation. Because that is not the kind of
% tutor I are describing, to avoid confusion, I use the term
% ``behavior'' instead.

@@

SMoL defines a semantic but not the syntax. However, to repsent SMoL
to students, I have to choose a syntax. I chose a Lispy syntax, which
is an artifact of the major population where I conducted studies: a
course (\Cref{s:populations}) using Racket
\citep{friedmanEssentialsProgrammingLanguages2001,
krishnamurthiProgrammingLanguagesApplication2007}. However, the Lispy
syntax also proved to be pedagogically valuable. I have found the
parentheses useful when discussing scope in conjunction with
local-binding features like \lstinline|let|. This avoids the various
confusing ``variable lifting'' semantics found in languages like
Python (\cite{politzPythonFullMonty2013}; see also posts like
\cite{froadieWhatScopeVariable2022}), where the actual defined range
of a variable is not apparent from the source code.

Nevertheless, most SMoL programs are easy to translate to other
languages. I present machine-translated Python and JavaScript versions
of my programs in \supplRef{Translated\_Programs.html}. Furthermore, I
intend to make a multi-lingual tutor in the future
(\Cref{s:future-work}).

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\citep{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my tutor (\Cref{s:tutor}).

Many tutoring systems focus on teaching programming (such as the
well-known and heavily studied LISP Tutor
\citep{andersonLISPTutor1985}), and in the process undoubtedly address
some program behavior misconceptions. My tutor differs in a notable
way: it does not try to teach programming per se. Instead, it assumes
a basic programming background and focuses entirely on program
behavior misconceptions and correcting them. I are not aware of a
tutoring system (in computer science) that has this specific design.

@@

I have another source of misconceptions. I have been creating a tutor
to answer my RQ2, which I explain in detail in \Cref{s:plan-rq2}. The
tutor includes many tasks that ask students to predict the result(s)
of running a program. Although these programs originate from
student-written programs (\Cref{s:find-misconceptions-with-quizius}), they
have been modified to fit tutoring purposes. Perhaps because of the
modification or because of the population variance, I have found
several unanticipated patterns of errors
(\Cref{s:misconceptions-tutor}), which suggest unknown misconceptions.

I plan to change the tutor such that it collects data to uncover these
unknown misconceptions and confirm known ones.

@@

Eventually, I expect to build a table of misconceptions. Each
misconception comes with multiple programs. For each of these
programs, the wrong result that corresponds to the misconception should
be commonly predicted by students from various population.
Furthermore, these students should give similar explanations for all
the wrong answers. For example, I might find a misconception called
\miscon{DefByRef} (See
\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
in \Cref{s:misconceptions-tutor} for a list of misconceptions that I
have found): when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.

@@

To answer this RQ, I plan to create a tutor (\Cref{s:design-tutor}) and
argue that the tutor is effective (\Cref{s:evaluation}).

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}).
My current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a tutor that
incorporates all these ideas (\Cref{s:tutor}) and have collected data
from all populations.

I would like to first explain key aspects of the tutor
(\Cref{s:tutorials,s:interpreting-tasks}) and then my plan to
incorporate ideas of refutation text (\Cref{s:refutationt-text}), case
comparison (\Cref{s:case-comparison}), and notional machines
(\Cref{s:notional-machine}). Finally, I present my plan to evaluate
the effectiveness of the tutor and how evaluation impacts the tutor
design (\Cref{s:evaluation}).

@@

There are many misconceptions to correct. To make the tutoring process
manageable, I have designed the tutor as a sequence of tutorials.
Inspired by prior research on language levels
(\Cref{s:pedagogic-techniques}), my sequence of tutorials is ordered,
and each tutorial covers one or more new language constructs
(\Cref{t:tutorials}). Each tutorial might be delivered as multiple modules to avoid forcing
students to continuously work for more than 30 minutes.

@@

The tutor is mostly consisted of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interpret
program. The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

Let's reconsider this program from \Cref{s:rq1-expected-results}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which corresponds
to the \miscon{DefByRef} misconception. If an MCQ only has few options
like this example, students have a great chance (in this case, 50\%)
of just guessing the correct answer or successfully using a process of
elimination. So I have been adding additional options that are not
obviously wrong. The tutor is adding many options, but with an ad hoc
rule. I plan to revise the rules of adding extra options.

\paragraph{Penalty to Wrong Choices}\label{s:penalty} To discourage
students from making a random choice, the tutor has been giving
penalty. Designing the strategy of assigning penalty is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because I do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item An alternative common strategy is to grade by completion.
  However, grading by completion alone encourages students to rush
  through the tutorials and hence might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grades as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the tutor.}; when a student fails a task, the tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

  \item Instead of multiple-choice, students must \emph{type} the
        answer into a text box. (The tutor normalizes text to accommodate
        variations.) This is intentional. First, I want to force
        reflection on the explanation just given, whereas with an MCQ,
        students could just guess. Second, I feel that students would
        find typing more onerous than clicking. In case students had just
        guessed on a question, my hope is that the penalty of having to
        type means, on \emph{subsequent} tasks, they would be more likely
        to pause and reflect before choosing.

@@

\paragraph{Refutation Text and Other Explanations}%
\label{s:refutationt-text}
The tutor has been presenting a refutation text when a student chooses
a wrong answer that corresponds to known misconception(s). Let's
reconsider the program in \Cref{s:interpreting-tasks}. If a student
answers \lstinline|0 0|, which suggests they hold the
\miscon{DefByRef} misconception, the tutor responds
\begin{quote}
  \lstinline|y| was bound to \lstinline|12| (i.e., the value of
  \lstinline|x|) rather than to \lstinline|x|. So changing the value
  of \lstinline|y| does not change the value of \lstinline|x|.
\end{quote}
If a student gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The first definition binds \lstinline|x| to \lstinline|12|. The
  second definition binds \lstinline|y| to the value of \lstinline|x|,
  which is \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}
If a student gave the correct answer, the tutor also presents the
generic explanation.

@@

The tutor provides opportunities for making case comparisons: each
interpreting task presents a SMoL program and the program's result.
These program-result pairs are concrete cases of the semantics of
SMoL. To incorporate the idea of case comparison, the tutor currently
prompts students to compare these cases and then to summarize the
rules of program behavior at the end of \emph{some} tutorials.

@@

Prior research argues (\Cref{s:pedagogic-techniques})that illustrating
how programs runs with notional machines (NMs) is beneficial for
students. I have been giving students opportunities and sometimes
explicitly require them to run programs in a NM visualization tool.

@@

To fully answer my research questions, I should also argue that my
tutor is effective at correcting misconceptions. In this section, I
describe how I want to refine the design of the tutor to help me make
the argument, what data I want to collect, and what (statistical)
analysis I want to perform on the data.

@@

Given this observation, I plan to set up the tutor such that
\begin{itemize}
  \item Every wrong answer corresponds to up to one misconception. In
        this case, I say the wrong answer \emph{represents} the
        misconception.
  \item Every misconception is represented by some questions.
  \item In each tutorial, a misconception is either not represented by
        any interpreting tasks or represented by at least two tasks.
  \item Within each tutorial, the order of interpreting tasks is
        randomized.
\end{itemize}
For each tutorial and each misconception, I plan to do a model
comparison (with the standard AIC and BIC criteria) on the following
models/theories:
\begin{description}
  \item[TaskSpecific] Some students are more likely to choose the
  representing wrong answers in certain tasks.
  \item[PositionSpecific] Students are more likely to choose the
  representing wrong answer in task that appears earlier (or later) in
  the tutorial.
\end{description}

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that I need further
investigation into the relationship between the tasks and
misconceptions. If the \textbf{PositionSpecific} model best explains
\emph{and} students perform better in later tasks, I deduce that the
tutor seems to be correcting misconceptions. If students perform
worse in later tasks, the tutor is potentially doing harm.

@@

The current tutor is not prepared for this analysis yet. I describe how
I plan to change the tutor in \Cref{s:future-work}.

@@

After running Quizius in the course (\Cref{s:populations}),
I took over as experts. Determining which is the right
answer is easy. Where expert knowledge is useful is in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

@@

The Quizius instrucment was set up by other people. I analyzed the
collected data.

@@

  \item I resolved ambiguities in some programs, either adding
        answer choices or even adding other questions to tease
        out different interpretations.

@@

I now examine what I learned from the tutor's interpreting tasks in
terms of program behavior misconceptions.

I iteratively created our final catalog of misconceptions (from the
perspective of the data in this paper). I started with misinterpreters
representing the misconceptions for which we
have reasonable validation (due to the prose in Quizius), so I call
them \emph{grounded} misconceptions. I then looked at wrong answers
not covered by these but chosen by students, and did my best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we identify with a $^\ddag$), which need to be
validated in the future. I then re-ran the misinterpreters against the
chosen answers. I terminated when all the remaining wrong answers were
either (a) found in very few students (we found a gap between 23\% and
13\%, and hence took 14\% as the threshold), (b) difficult for us to
attribute to a misconception, or (c) appeared to me to be ``Lispy''
and hence not of broad interest.

@@

This section describe the current status of the tutor.

@@

\Cref{f:smol-tutor-example} illustrates an interpreting task as
described in \Cref{s:interpreting-tasks}. The Tutor also asked students
to perform some other activities, like showing programs and asking for
their heap content, which I do not cover in this paper.

@@

In addition to asking students questions, the Tutor along the way
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. I therefore refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial. I plan to keep
goal sentences in the tutor.

@@

% Recall that the SMoL Tutor is not only a collection of MCQs: it is
% also a \emph{tutor}! So far, I have investigated the value of the
% MCQs. Now I examine its tutoring aspect. Concretely, I ask:
% \begin{center}

@@

This section presents part of my preliminary answer to \textbf{RQ2}
(\Cref{s:plan-rq2}): is the tutor effective?

@@

I also perform a logistic regression to see whether these
improvements are significant (at a $p < 0.05$ threshold); details are
in \supplRef{Paper.html}. Of the 11:
\begin{itemize}

@@

Overall, the data suggest that the Tutor \emph{most likely did not do
harm} and perhaps even \emph{may have done some good}. Revisions on
the tutor (\Cref{s:future-work}), in light of misinterpreters, is
necessarily to give a more accurate evaluation. For example I found
far fewer problems than I would have liked. Concretely, only 40 of the
71 eligible problems (after removing the non-SMoL modules) were useful
in the above analysis.

@@

I have learned (\Cref{s:pedagogic-techniques}) that refutation text,
case comparison, notional machines, and language levels are helpful. I
plan to explore more literature and improve the tutor if I find new
helpful techniques.

@@

The preliminary tutor provides a ``Copy This Program'' button that
helps students to run the program in DrRacket, where a SMoL
implementation is provided.

@@

The latest tutor provides a button on the top-right of each program.
If students click the button, they will open a new browser tab that
shows the NM populated with the program. I have deployed the latest
tutor to the latest population at Brown (i.e., University 1), but I
have not analyzed the data, so the new tutor is not presented in
the preliminary results.

@@

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
I found that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from the tutor, there is still a risk of having biases in student
errors. This semester (Fall 2023) I am already deploying a newer SMoL
Tutor at Brown. This new tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new tutor data with the current results,
i.e., a comparison between uni-syntax tutor and multi-syntax tutor.

@@

Another issue with the current design is that students might run a
program before giving an answer. I don't know how often this
happens. But, given that students receive a penalty (\Cref{s:penalty})
for giving wrong answers, they might do so when the program is overly
complicated. This problem is more of an issue after the switch to a
web-based notional machine because now the ``cost'' of running a
program is cheaper. My plan is to change the tutor so that it provides
a program-specific link \emph{only} after each interpreting task.

@@

The tutor currently prompts students to compare program-output pairs
and then to summarize the rules of program behavior at the end of
\emph{some} tutorials.

>>>>

% TODO
% - clean-up the background sections

@@

@@

  I propose a plan for finding misconceptions possibly not anticipated
  by experts, and the design of an automated, self-guided tutoring
  system. The Tutor builds on strategies in the cognitive and
  educational literature and is explicitly designed around identifying
  and correcting misconceptions. Preliminary results suggest (a) the
  misconceptions I found are widespread, and (b) the Tutor appears to
  improve understanding.
\end{abstract}

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions (e.g.,
Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python
\citep{bernhardtWat2012, guhaEssenceJavascript2010,
politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still,
this semantic core bridges many syntaxes, and understanding it helps
when transferring knowledge from old languages to new ones. In
recognition of this deep commonality, in this proposal I choose to call
this the \emph{Standard Model of Languages} (SMoL).

@@

This led to my \textbf{research questions}:

@@

@@

SMoL defines a semantic but not the syntax. However, to represent SMoL
to students, I have to choose a syntax. I chose a Lispy syntax, which
is an artifact of the major population where I conducted studies: a
course (\Cref{s:populations}) using Racket
\citep{friedmanEssentialsProgrammingLanguages2001,
krishnamurthiProgrammingLanguagesApplication2007}. However, the Lispy
syntax also proved to be pedagogically valuable. I have found the
parentheses useful when discussing scope in conjunction with
local-binding features like \lstinline|let|. This avoids the various
confusing ``variable lifting'' semantics found in languages like
Python (\cite{politzPythonFullMonty2013}; see also posts like
\cite{froadieWhatScopeVariable2022}), where the actual defined range
of a variable is not apparent from the source code.

Nevertheless, most SMoL programs are easy to translate to other
languages. Furthermore, I intend to make a multi-lingual Tutor in the
future (\Cref{s:future-work}).

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\citep{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my Tutor (\Cref{s:tutor}).

Many tutoring systems focus on teaching programming (such as the
well-known and heavily studied LISP Tutor
\citep{andersonLISPTutor1985}), and in the process undoubtedly address
some program behavior misconceptions. My Tutor differs in a notable
way: it does not try to teach programming per se. Instead, it assumes
a basic programming background and focuses entirely on program
behavior misconceptions and correcting them. I am not aware of a
tutoring system (in computer science) that has this specific design.

@@

I have another source of misconceptions. I have been creating a Tutor
to answer my RQ2, which I explain in detail in \Cref{s:plan-rq2}. The
Tutor includes many tasks that ask students to predict the result(s)
of running a program. Although these programs originate from
student-written programs (\Cref{s:find-misconceptions-with-quizius}), they
have been modified to fit tutoring purposes. Perhaps because of the
modification or because of the population variance, I have found
several unanticipated patterns of errors
(\Cref{s:misconceptions-tutor}), which suggest unknown misconceptions.

I plan to change the Tutor so that it collects data to uncover these
unknown misconceptions and confirm known ones.

@@

Eventually, I expect to build a table of misconceptions. Each
misconception comes with multiple programs. For each of these
programs, the wrong result that corresponds to the misconception should
be commonly predicted by students from various populations.
Furthermore, these students should give similar explanations for all
the wrong answers. For example, I might find a misconception called
\miscon{DefByRef} (See
\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
in \Cref{s:misconceptions-tutor} for a list of misconceptions that I
have found): when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.

@@

To answer this RQ, I plan to create a Tutor (\Cref{s:design-tutor}) and
argue that the Tutor is effective (\Cref{s:evaluation}).

@@

To answer this RQ, I have done some literature review on what
pedagogic techniques are helpful (\Cref{s:pedagogic-techniques}).
My current takeaway is that refutation text, case comparison, notional
machines, and language levels are helpful. I have created a Tutor that
incorporates all these ideas (\Cref{s:tutor}) and have collected data
from all populations.

I would like to first explain key aspects of the Tutor
(\Cref{s:tutorials,s:interpreting-tasks}) and then my plan to
incorporate ideas of refutation text (\Cref{s:refutationt-text}), case
comparison (\Cref{s:case-comparison}), and notional machines
(\Cref{s:notional-machine}). Finally, I present my plan to evaluate
the effectiveness of the Tutor and how evaluation impacts the Tutor
design (\Cref{s:evaluation}).

@@

There are many misconceptions to correct. To make the tutoring process
manageable, I have designed the Tutor as a sequence of tutorials.
Inspired by prior research on language levels
(\Cref{s:pedagogic-techniques}), my sequence of tutorials is ordered,
and each tutorial covers one or more new language constructs
(\Cref{t:tutorials}). Each tutorial might be delivered as multiple modules to avoid forcing
students to continuously work for more than 30 minutes.

@@

The Tutor mostly consists of \emph{interpreting tasks}:
multiple-choice questions (MCQs) that asks students to interpret
program. The options include (among other options
that I will explain shortly) the correct answer and wrong answers that
correspond to known misconceptions. Because students might conceive
wrong answers that I do not anticipate, the Tutor has been providing
an ``Other'' option that, once chosen, allows students to enter
arbitrary answers.

Let's reconsider this program from \Cref{s:rq1-expected-results}
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
In this case, the options should include the correct answer
\lstinline|12 0| and a wrong answer \lstinline|0 0|, which corresponds
to the \miscon{DefByRef} misconception. If an MCQ only has few options
like this example, students have a great chance (in this case, 50\%)
of just guessing the correct answer or successfully using a process of
elimination. So I have been adding additional options that are not
obviously wrong. The Tutor is adding many options, but with an ad hoc
rule. I plan to revise the rules of adding extra options.

\paragraph{Penalty to Wrong Choices}\label{s:penalty} To discourage
students from making a random choice, the Tutor has been giving
penalties. Designing the strategy of assigning penalties is non-trivial:
\begin{itemize}
  \item A common strategy is to grade students by their correctness.
  However, grading by correctness is not suitable for a tutoring
  system because I do not mind students doing badly on earlier tasks
  as long as they eventually master the concepts. Actually, if
  students generally succeed in all tasks, the tutoring system is
  failing -- it does not teach anything to the students.
  \item An alternative common strategy is to grade by completion.
  However, grading by completion alone encourages students to rush
  through the tutorials and, hence, might encourage students to make
  random choices.
\end{itemize}
In short, I want to make students feel as follows:
\begin{itemize}
  \item They won't lose grades as long as they finish the tutorials.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I have been doing the following: when I
deployed the Tutor in my institution, University 1, I graded by
completion\footnote{I encourage collaborators to grade by
completeness, but I have no real control over how they actually use
the Tutor.}; when a student fails a task, the Tutor gives an
\emph{extra} task:
\begin{enumerate}

@@

  \item Instead of multiple-choice, students must \emph{type} the
        answer into a text box. (The Tutor normalizes text to accommodate
        variations.) This is intentional. First, I want to force
        reflection on the explanation just given, whereas with an MCQ,
        students could just guess. Second, I feel that students would
        find typing more onerous than clicking. In case students had just
        guessed on a question, my hope is that the penalty of having to
        type means, on \emph{subsequent} tasks, they would be more likely
        to pause and reflect before choosing.

@@

\paragraph{Refutation Text and Other Explanations}%
\label{s:refutationt-text}
The Tutor has been presenting a refutation text when a student chooses
a wrong answer that corresponds to known misconception(s). Let's
reconsider the program in \Cref{s:interpreting-tasks}. If a student
answers \lstinline|0 0|, which suggests they hold the
\miscon{DefByRef} misconception, the Tutor responds
\begin{quote}
  \lstinline|y| was bound to \lstinline|12| (i.e., the value of
  \lstinline|x|) rather than to \lstinline|x|. So changing the value
  of \lstinline|y| does not change the value of \lstinline|x|.
\end{quote}
If a student gives a wrong answer that does not correspond to a
misconception (either by choosing one of the generated options or by
choosing the ``Other'' option), the Tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
Tutor responds
\begin{quote}
  The first definition binds \lstinline|x| to \lstinline|12|. The
  second definition binds \lstinline|y| to the value of \lstinline|x|,
  which is \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}
If a student gives the correct answer, the Tutor also presents the
generic explanation.

@@

The Tutor provides opportunities for making case comparisons: each
interpreting task presents a SMoL program and the program's result.
These program-result pairs are concrete cases of the semantics of
SMoL. To incorporate the idea of case comparison, the Tutor currently
prompts students to compare these cases and then to summarize the
rules of program behavior at the end of \emph{some} tutorials.

@@

Prior research argues (\Cref{s:pedagogic-techniques})that illustrating
how programs run with notional machines (NMs) is beneficial for
students. I have been giving students opportunities and sometimes
explicitly requiring them to run programs in the NM visualization tool.

@@

To fully answer my research questions, I should also argue that the
Tutor is effective at correcting misconceptions. In this section, I
describe how I want to refine the design of the Tutor to help me make
the argument, what data I want to collect, and what (statistical)
analysis I want to perform on the data.

@@

Given this observation, I plan to set up the Tutor such that
\begin{itemize}
  \item Every wrong answer corresponds to up to one misconception. In
        this case, I say the wrong answer \emph{represents} the
        misconception.
  \item Every misconception is represented by some questions.
  \item In each tutorial, a misconception is either not represented by
        any interpreting tasks or represented by at least two tasks.
  \item Within each tutorial, the order of interpreting tasks is
        randomized.
\end{itemize}
For each tutorial and each misconception, I plan to do a model
comparison (with the standard AIC and BIC criteria) on the following
models/theories:
\begin{description}
  \item[TaskSpecific] Some students are more likely to choose the
  representing wrong answers in certain tasks.
  \item[PositionSpecific] Students are more likely to choose the
  representing wrong answer in task that appears earlier (or later) in
  the tutorial.
\end{description}

If the \textbf{TaskSpecific} model best explains the data, I deduce
that some programs are more confusing and that I need further
investigation into the relationship between the tasks and
misconceptions. If the \textbf{PositionSpecific} model best explains
\emph{and} students perform better in later tasks, I deduce that the
Tutor seems to be correcting misconceptions. If students perform
worse in later tasks, the Tutor is potentially doing harm.

@@

The current Tutor is not prepared for this analysis yet. I describe how
I plan to change the Tutor in \Cref{s:future-work}.

@@

After running Quizius in the course (\Cref{s:populations}),
I took over as experts. Determining which is the right
answer is easy, whereas expert knowledge is useful is in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

@@

The Quizius instrument was set up by other people. I analyzed the
collected data.

@@

  \item I resolved ambiguities in some programs by adding
        answer choices or even other questions to tease
        out different interpretations.

@@

I now examine what I learned from the Tutor's interpreting tasks in
terms of program behavior misconceptions.

I iteratively created our final catalog of misconceptions (from the
perspective of the data in this document). I started with misinterpreters
representing the misconceptions for which we
have reasonable validation (due to the prose in Quizius), so I call
them \emph{grounded} misconceptions. I then looked at wrong answers
not covered by these but chosen by students, and did my best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we identify with a $^\ddag$), which need to be
validated in the future. I then re-ran the misinterpreters against the
chosen answers. I terminated when all the remaining wrong answers were
either (a) found in very few students (we found a gap between 23\% and
13\%, and hence took 14\% as the threshold), (b) difficult for us to
attribute to a misconception, or (c) appeared to me to be ``Lispy''
and hence not of broad interest.

@@

This section describe the current status of the Tutor.

@@

\Cref{f:smol-tutor-example} illustrates an interpreting task as
described in \Cref{s:interpreting-tasks}. The Tutor also asked students
to perform some other activities, like showing programs and asking for
their heap content, which I do not cover in this document.

@@

In addition to asking students questions, the Tutor along the way
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. I therefore refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial. I plan to keep
goal sentences in the Tutor.

@@

% Recall that the SMoL Tutor is not only a collection of MCQs: it is
% also a \emph{Tutor}! So far, I have investigated the value of the
% MCQs. Now I examine its tutoring aspect. Concretely, I ask:
% \begin{center}

@@

This section presents part of my preliminary answer to \textbf{RQ2}
(\Cref{s:plan-rq2}): is the Tutor effective?

@@

I also perform a logistic regression to see whether these
improvements are significant (at a $p < 0.05$ threshold). Of the 11:
\begin{itemize}

@@

Overall, the data suggest that the Tutor \emph{most likely did not do
harm} and perhaps even \emph{may have done some good}. Revisions on
the Tutor (\Cref{s:future-work}), in light of misinterpreters, is
necessarily to give a more accurate evaluation. For example I found
far fewer problems than I would have liked. Concretely, only 40 of the
71 eligible problems (after removing the non-SMoL modules) were useful
in the above analysis.

@@

I have learned (\Cref{s:pedagogic-techniques}) that refutation text,
case comparison, notional machines, and language levels are helpful. I
plan to explore more literature and improve the Tutor if I find new
helpful techniques.

@@

The preliminary Tutor provides a ``Copy This Program'' button that
helps students to run the program in DrRacket, where a SMoL
implementation is provided.

@@

The latest Tutor provides a button on the top-right of each program.
If students click the button, they will open a new browser tab that
shows the NM populated with the program. I have deployed the latest
Tutor to the latest population at Brown (i.e., University 1), but I
have not analyzed the data, so the new Tutor is not presented in
the preliminary results.

@@

The processes described in
\Cref{s:find-misconceptions-with-quizius,s:find-misconceptions-with-tutor}
have been conducted with SMoL's Lispy syntax (\Cref{s:smol}). This can
potentially introduce biases to the found misconceptions. For example,
I found that some students think the following program
produces \lstinline|'(4 5)| rather than \lstinline|'(1 2)|.
\begin{lstlisting}
  (filter (lambda (n) (> 3 n)) '(1 2 3 4 5))
\end{lstlisting}
This checks whether $\verb|3| > \verb|n|$, but those students
presumably vocalized it as ``greater than \lstinline|3|,
\lstinline|n|?''. Although I have tried to remove programs like this
from the Tutor, there is still a risk of having biases in student
errors. This semester (Fall 2023) I am already deploying a newer SMoL
Tutor at Brown. This new Tutor is able to display programs in
JavaScript and Python, in addition to the Lispy syntax. I plan to
compare in Spring 2024 the new Tutor data with the current results,
i.e., a comparison between uni-syntax Tutor and multi-syntax Tutor.

@@

Another issue with the current design is that students might run a
program before giving an answer. I don't know how often this
happens. But, given that students receive a penalty (\Cref{s:penalty})
for giving wrong answers, they might do so when the program is overly
complicated. This problem is more of an issue after the switch to a
web-based notional machine because now the ``cost'' of running a
program is cheaper. My plan is to change the Tutor so that it provides
a program-specific link \emph{only} after each interpreting task.

@@

The Tutor currently prompts students to compare program-output pairs
and then to summarize the rules of program behavior at the end of
\emph{some} tutorials.
